---
title: "Modelling and Forecasting Spanish Real GDP Growth"
subtitle: "Econometrics 20203 -- Group Assignment, Part II"
author: "Stefano Graziosi"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-tools: true
    theme: cosmo
    fig-align: center
    fig-width: 9
    fig-height: 5
    tbl-cap-location: top
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
    library(dplyr)
    library(tidyr)
    library(ggplot2)
    library(scales)
    library(tseries)
    library(forecast)
    library(vars)
    library(BVAR)
    library(strucchange)
    library(urca)
    library(zoo)
    library(knitr)
    library(kableExtra)
    library(gridExtra)
    library(RColorBrewer)
})

# Project palette — used consistently across plots
pal_main <- c(
    GDP   = "#1f4e79",
    INF   = "#c0504d",
    RATE  = "#7030a0",
    AUX1  = "#2c8a4f",
    AUX2  = "#e8a33d"
)
pal_models <- c(
    "Actual"               = "#111111",
    "BVAR(2) -- preferred" = "#c0504d",
    "VAR(2)"               = "#1f4e79",
    "AR(2)"                = "#2c8a4f",
    "ARIMA (BIC)"          = "#7030a0"
)

# Consistent ggplot theme
theme_pub <- function(base_size = 11) {
    theme_minimal(base_size = base_size) +
        theme(
            plot.title       = element_text(face = "bold", size = rel(1.05)),
            plot.subtitle    = element_text(colour = "grey25", size = rel(0.95)),
            plot.caption     = element_text(colour = "grey45", size = rel(0.80),
                                            hjust = 0),
            axis.title       = element_text(colour = "grey20"),
            strip.text       = element_text(face = "bold", colour = "grey15"),
            panel.grid.minor = element_blank(),
            panel.grid.major = element_line(colour = "grey92", linewidth = 0.3),
            legend.position  = "bottom",
            legend.title     = element_text(face = "bold", size = rel(0.85)),
            legend.text      = element_text(size = rel(0.85)),
            legend.key.width = grid::unit(1.2, "lines")
        )
}
theme_set(theme_pub())

# Reproducibility
set.seed(20203)

# Helper for kable tables
ftab <- function(x, digits = 3, caption = NULL, ...) {
    knitr::kable(x, digits = digits, caption = caption, booktabs = TRUE, ...) |>
        kable_styling(
            bootstrap_options = c("striped", "hover", "condensed"),
            full_width = FALSE, position = "center"
        )
}
```

# Introduction

This report specifies and evaluates an econometric model for the year-on-year real
GDP growth rate of **Spain**. Following the structure required by the assignment, the
analysis proceeds in six steps: (i) descriptive analysis of the variables;
(ii) a discussion of alternative specifications; (iii) diagnostic checks on the
preferred model; (iv) tests of economic hypotheses on the model parameters;
(v) a one-step-ahead pseudo-out-of-sample forecasting exercise over the last
ten years of the sample; and (vi) a comparison with the AR(2) and VAR(2) benchmarks
prescribed by the brief, where the VAR also includes year-on-year inflation
and the short-term interest rate.

We work with annual observations from the **Global Macroeconomic Database**
(Müller, Schularick, and Trebesch, 2024), which provides harmonised long
series for real GDP, the consumer-price-based inflation rate, and a
short-term interest rate. We restrict the sample to **1960--2025** to focus on the
modern, post-Bretton-Woods Spanish economy: this avoids the structural breaks
associated with the two World Wars, the Spanish Civil War (1936--39), and Franco-era
autarky (1939--59), which would invalidate the assumption of a stable
data-generating process underpinning standard time-series inference. The
resulting sample provides 65 year-on-year growth observations, sufficient
for the AR(2), ARIMA, VAR(2), and BVAR(2) models considered below.

# Data

```{r data-preparation}
gmd <- read.csv("../data/GMD.csv")

spain_full <- gmd |>
    dplyr::filter(ISO3 == "ESP") |>
    dplyr::arrange(year) |>
    dplyr::select(year, rGDP, infl, strate) |>
    dplyr::filter(!is.na(rGDP), !is.na(infl), !is.na(strate)) |>
    dplyr::mutate(gdp_growth = (rGDP / dplyr::lag(rGDP) - 1) * 100) |>
    dplyr::filter(!is.na(gdp_growth))

# Modern subsample for main analysis
spain <- spain_full |> dplyr::filter(year >= 1961)

start_year <- min(spain$year)
end_year <- max(spain$year)

ts_gdp <- ts(spain$gdp_growth, start = start_year, frequency = 1)
ts_infl <- ts(spain$infl, start = start_year, frequency = 1)
ts_strate <- ts(spain$strate, start = start_year, frequency = 1)

var_data <- cbind(
    GDP_Growth = ts_gdp,
    Inflation = ts_infl,
    Interest_Rate = ts_strate
)
```

The working sample covers `r start_year`--`r end_year` (`r nrow(spain)` annual
observations). The three variables are: year-on-year real GDP growth
($\Delta y_t = 100 \cdot (Y_t / Y_{t-1} - 1)$, percent), CPI inflation
($\pi_t$, percent), and the short-term nominal interest rate ($i_t$, percent).

# Descriptive Analysis

## Time-series plots

```{r descriptive-plots, fig.height=6.5}
plot_df <- spain |>
    dplyr::select(year, gdp_growth, infl, strate) |>
    tidyr::pivot_longer(-year, names_to = "series", values_to = "value") |>
    dplyr::mutate(series = factor(series,
        levels = c("gdp_growth", "infl", "strate"),
        labels = c(
            "Real GDP growth (%, y/y)",
            "CPI inflation (%, y/y)",
            "Short-term interest rate (%)"
        )
    ))

# Mark salient macro events for visual context
events <- tibble::tibble(
    year  = c(1973, 1986, 1999, 2008, 2020),
    label = c("Oil shock", "EEC entry", "Euro", "GFC", "COVID-19")
)

series_pal <- c(
    "Real GDP growth (%, y/y)"     = unname(pal_main["GDP"]),
    "CPI inflation (%, y/y)"       = unname(pal_main["INF"]),
    "Short-term interest rate (%)" = unname(pal_main["RATE"])
)

ggplot(plot_df, aes(year, value, colour = series, fill = series)) +
    geom_hline(yintercept = 0, colour = "grey55", linetype = "dashed",
               linewidth = 0.35) +
    geom_vline(
        data = events, aes(xintercept = year),
        colour = "grey78", linetype = "dotted", linewidth = 0.4,
        inherit.aes = FALSE
    ) +
    geom_text(
        data = events, aes(x = year, y = Inf, label = label),
        angle = 90, hjust = 1.05, vjust = -0.3, size = 2.7, colour = "grey35",
        inherit.aes = FALSE
    ) +
    geom_area(alpha = 0.10, colour = NA) +
    geom_line(linewidth = 0.75) +
    facet_wrap(~series, ncol = 1, scales = "free_y") +
    scale_colour_manual(values = series_pal, guide = "none") +
    scale_fill_manual(values = series_pal, guide = "none") +
    scale_x_continuous(breaks = seq(1960, 2025, 10),
                       expand = expansion(mult = c(0.01, 0.01))) +
    labs(
        title    = "Spain: macroeconomic time series, 1961--2025",
        subtitle = "Annual frequency; vertical guides mark salient macro events",
        caption  = "Source: Müller, Schularick & Trebesch (2024), Global Macro Database.",
        x = NULL, y = "Percent"
    )
```

The series display the expected post-war Spanish business-cycle features. GDP
growth is centered around three percent on average but with two pronounced
contractions: the global financial crisis (2008--09 and the subsequent euro-area
sovereign-debt crisis of 2011--13) and the COVID-19 collapse of 2020 (--10.9 percent).
Inflation rose sharply in the 1970s and gradually moderated after EEC accession
(1986) and especially after the adoption of the euro (1999); the post-2021
energy shock briefly pushed it back above eight percent. The short rate broadly
tracks the inflation pattern, declining steadily from double digits in the 1980s
to the negative territory imposed by the ECB during 2014--21 and rising again in 2023--25.

## Summary statistics

```{r summary-stats}
desc <- spain |>
    dplyr::select(
        `GDP growth (%)` = gdp_growth,
        `Inflation (%)` = infl,
        `Short rate (%)` = strate
    ) |>
    tidyr::pivot_longer(everything(), names_to = "Variable", values_to = "x") |>
    dplyr::group_by(Variable) |>
    dplyr::summarise(
        N = dplyr::n(),
        Mean = mean(x),
        `Std. dev.` = sd(x),
        Min = min(x),
        Median = median(x),
        Max = max(x),
        `Skewness` = mean((x - mean(x))^3) / sd(x)^3,
        `Kurtosis` = mean((x - mean(x))^4) / sd(x)^4,
        .groups = "drop"
    )
ftab(desc, digits = 2, caption = "Summary statistics, 1961--2025")
```

GDP growth has heavy left-tail behaviour (negative skew, excess kurtosis above
three) driven by the GFC and COVID outliers. Inflation and the short rate are
right-skewed, reflecting the high-inflation episodes of the 1970s and early 1980s.

## Marginal distributions

```{r distribution-plots, fig.height=4.2}
dist_df <- spain |>
    dplyr::select(gdp_growth, infl, strate) |>
    tidyr::pivot_longer(everything(), names_to = "series", values_to = "value") |>
    dplyr::mutate(series = factor(series,
        levels = c("gdp_growth", "infl", "strate"),
        labels = c("GDP growth", "Inflation", "Short rate")
    ))

mean_df <- dist_df |>
    dplyr::group_by(series) |>
    dplyr::summarise(mean = mean(value), median = median(value), .groups = "drop")

p_hist <- ggplot(dist_df, aes(value, fill = series, colour = series)) +
    geom_histogram(aes(y = after_stat(density)), bins = 22,
                   alpha = 0.55, colour = "white", linewidth = 0.25) +
    geom_density(alpha = 0, linewidth = 0.7) +
    geom_vline(data = mean_df, aes(xintercept = mean, colour = series),
               linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
    facet_wrap(~series, scales = "free", nrow = 1) +
    scale_fill_manual(values = unname(pal_main[1:3]), guide = "none") +
    scale_colour_manual(values = unname(pal_main[1:3]), guide = "none") +
    labs(title = "Empirical density and histogram",
         subtitle = "Dashed line = sample mean",
         x = "Percent", y = "Density")

# Theoretical Q-Q plots vs Normal
p_qq <- ggplot(dist_df, aes(sample = value, colour = series)) +
    stat_qq(size = 1.2, alpha = 0.85) +
    stat_qq_line(linewidth = 0.45, linetype = "dashed", colour = "grey40") +
    facet_wrap(~series, scales = "free", nrow = 1) +
    scale_colour_manual(values = unname(pal_main[1:3]), guide = "none") +
    labs(title = "Normal Q--Q plots",
         subtitle = "Departures from the dashed line indicate non-normality",
         x = "Theoretical quantile", y = "Sample quantile")

gridExtra::grid.arrange(p_hist, p_qq, nrow = 2)
```

The histograms confirm the asymmetries already revealed by skewness and kurtosis:
GDP growth has a long, thin left tail (the GFC and COVID episodes) and a
right-skewed inflation distribution dominated by the 1970s. The Q--Q plots
display systematic deviations from the dashed 45-degree line, particularly in
the tails — formal tests of normality on the model residuals are reported in
the diagnostic battery below.

## Bivariate structure

```{r scatter-corr, fig.height=4.2}
pair_df <- spain |>
    dplyr::select(gdp_growth, infl, strate)

cor_mat <- round(cor(pair_df), 2)
cor_df <- as.data.frame(as.table(cor_mat)) |>
    setNames(c("Var1", "Var2", "Correlation"))

p_corr <- ggplot(cor_df, aes(Var1, Var2, fill = Correlation)) +
    geom_tile(colour = "white", linewidth = 0.4) +
    geom_text(aes(label = sprintf("%.2f", Correlation)),
              size = 3.6, colour = "grey15") +
    scale_fill_gradient2(low = "#3a76b8", mid = "grey95", high = "#c0504d",
                        midpoint = 0, limits = c(-1, 1)) +
    scale_x_discrete(labels = c("GDP growth", "Inflation", "Short rate")) +
    scale_y_discrete(labels = c("GDP growth", "Inflation", "Short rate")) +
    coord_fixed() +
    labs(title = "Contemporaneous correlation matrix",
         x = NULL, y = NULL, fill = "ρ")

# Cross-correlation function: GDP growth vs short rate
ccf_obj <- ccf(ts_strate, ts_gdp, lag.max = 8, plot = FALSE)
ccf_df <- data.frame(lag = as.numeric(ccf_obj$lag),
                     acf = as.numeric(ccf_obj$acf))
crit <- 1.96 / sqrt(length(ts_gdp))

p_ccf <- ggplot(ccf_df, aes(lag, acf)) +
    geom_hline(yintercept = 0, colour = "grey50") +
    geom_hline(yintercept = c(-crit, crit), colour = "#c0504d",
               linetype = "dashed", linewidth = 0.4) +
    geom_segment(aes(xend = lag, yend = 0), colour = unname(pal_main["GDP"]),
                 linewidth = 0.9) +
    geom_point(colour = unname(pal_main["GDP"]), size = 1.6) +
    labs(title = "CCF: short rate (lead) vs. GDP growth",
         subtitle = "Negative lag k: short rate leads GDP growth by k years",
         x = "Lag (years)", y = "Cross-correlation")

gridExtra::grid.arrange(p_corr, p_ccf, nrow = 1, widths = c(0.45, 0.55))
```

The contemporaneous correlation between GDP growth and the short rate is
mildly negative, while inflation and the short rate are strongly positively
correlated — the latter consistent with a Taylor-rule-style policy reaction.
The cross-correlation function (right panel) shows that short-rate movements
lead GDP growth at horizons of 1--3 years, with a meaningful negative
lead-correlation, suggesting a monetary-transmission channel that we revisit
when discussing impulse responses.

## Rolling-window moments

```{r rolling-stats, fig.height=4.2}
window <- 7
roll_df <- spain |>
    dplyr::transmute(
        year,
        gdp_mean   = zoo::rollapply(gdp_growth, window, mean,
                                    fill = NA, align = "right"),
        gdp_sd     = zoo::rollapply(gdp_growth, window, sd,
                                    fill = NA, align = "right"),
        infl_mean  = zoo::rollapply(infl, window, mean,
                                    fill = NA, align = "right"),
        rate_mean  = zoo::rollapply(strate, window, mean,
                                    fill = NA, align = "right")
    )

p_rmean <- roll_df |>
    dplyr::select(year, GDP = gdp_mean, Inflation = infl_mean, Rate = rate_mean) |>
    tidyr::pivot_longer(-year, names_to = "series") |>
    ggplot(aes(year, value, colour = series)) +
        geom_hline(yintercept = 0, colour = "grey60", linetype = "dashed",
                   linewidth = 0.35) +
        geom_line(linewidth = 0.8) +
        scale_colour_manual(values = c(GDP = unname(pal_main["GDP"]),
                                       Inflation = unname(pal_main["INF"]),
                                       Rate = unname(pal_main["RATE"])),
                            name = NULL) +
        labs(title = paste0(window, "-year rolling means"),
             subtitle = "Tracks low-frequency level shifts",
             x = NULL, y = "Percent")

p_rsd <- ggplot(roll_df, aes(year, gdp_sd)) +
    geom_line(colour = unname(pal_main["GDP"]), linewidth = 0.85) +
    geom_point(colour = unname(pal_main["GDP"]), size = 1.3) +
    labs(title = paste0(window, "-year rolling std. dev. (GDP growth)"),
         subtitle = "Tracks volatility — note the spike around 2008--2020",
         x = NULL, y = "Std. deviation (pp)")

gridExtra::grid.arrange(p_rmean, p_rsd, nrow = 1, widths = c(0.55, 0.45))
```

The rolling means highlight the secular decline of inflation and the short
rate from the high levels of the 1970s and early 1980s to near-zero readings
in the post-2014 ECB unconventional-policy window, while average GDP growth
shifts down at the time of the GFC. The rolling standard deviation of growth
is roughly flat at ~2 pp until 2007, then jumps as the GFC and COVID episodes
enter the window — direct evidence of time-varying second moments and a
preview of the structural-stability tests reported below.

## Autocorrelation structure

```{r acf-pacf, fig.height=4.2}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
Acf(ts_gdp, main = "ACF: GDP growth", lag.max = 16)
Pacf(ts_gdp, main = "PACF: GDP growth", lag.max = 16)
```

The autocorrelation function decays geometrically from a first-order coefficient
of approximately 0.45, while the partial autocorrelation function cuts off after
the first lag. Both patterns are consistent with a stationary low-order
autoregressive process and motivate the AR(p) and ARIMA candidates considered below.

## Stationarity tests

Tests of the unit-root hypothesis are reported in @tbl-unit-roots. We combine
the Augmented Dickey--Fuller test (null: unit root) with the KPSS test
(null: level stationarity) so that conclusions agree only when the two procedures
both reject in the appropriate direction.

```{r}
#| label: tbl-unit-roots
#| tbl-cap: "Unit-root tests, 1961--2025"
adf <- function(x) suppressWarnings(tseries::adf.test(x, alternative = "stationary"))
kpss <- function(x) suppressWarnings(tseries::kpss.test(x, null = "Level"))

ur_tab <- data.frame(
    Variable = c("GDP growth", "Inflation", "Short rate"),
    `ADF stat` = c(adf(ts_gdp)$statistic, adf(ts_infl)$statistic, adf(ts_strate)$statistic),
    `ADF p-value` = c(adf(ts_gdp)$p.value, adf(ts_infl)$p.value, adf(ts_strate)$p.value),
    `KPSS stat` = c(kpss(ts_gdp)$statistic, kpss(ts_infl)$statistic, kpss(ts_strate)$statistic),
    `KPSS p-value` = c(kpss(ts_gdp)$p.value, kpss(ts_infl)$p.value, kpss(ts_strate)$p.value),
    check.names = FALSE
)
ftab(ur_tab, digits = 3)
```

GDP growth is stationary on visual inspection and on the KPSS test (high p-value);
the ADF p-value of about 0.12 reflects the modest power of the test in samples
of this size, but combined with the clearly mean-reverting behaviour and the
ACF/PACF pattern we treat $\Delta y_t$ as $I(0)$. Inflation and the short rate
display higher persistence; the KPSS test rejects level stationarity at 5%
for both. Following the macroeconomic literature, we keep the variables in
levels rather than differencing them, since (a) the Bayesian VAR's Minnesota
prior shrinks coefficients toward a random walk and is robust to persistent
regressors, and (b) the short interest rate and inflation are theoretically
bounded and cointegration with growth is plausible. The diagnostic and
forecasting results are insensitive to these choices.

# Structural Stability and Breakpoint Tests

The descriptive evidence — the secular decline in inflation and short-term
rates, the level shift in the rolling mean of GDP growth, and the spike in
its rolling volatility — points to potential structural change in the joint
data-generating process. We assess this systematically with (i) Chow $F$-tests
at four economically motivated candidate dates, (ii) the Quandt--Andrews
sup-$F$ test, which leaves the break date *unknown*, and (iii) the
Bai--Perron multiple-breakpoint procedure. The CUSUM test of recursive
residuals is reported as a complementary visual diagnostic.

```{r chow-prep}
# Build the AR(2) regression frame used by all break tests
n_full <- nrow(spain)
df_ar <- data.frame(
    year = spain$year[3:n_full],
    y    = spain$gdp_growth[3:n_full],
    y1   = spain$gdp_growth[2:(n_full - 1)],
    y2   = spain$gdp_growth[1:(n_full - 2)]
)
```

## Chow tests at four candidate breakpoints

The four candidates were chosen *a priori* from the historical record of
post-1960 Spanish macro shocks:

- **CH1 (1973)** — first OPEC oil shock and the end of Bretton Woods.
- **CH2 (1986)** — Spain's accession to the European Economic Community.
- **CH3 (2008)** — onset of the Global Financial Crisis.
- **CH4 (2020)** — COVID-19 pandemic shock.

Under $H_0$ the AR(2) coefficients $(c, \phi_1, \phi_2)$ are constant across
the two sub-samples defined by the candidate date.

```{r chow-tests}
chow_at <- function(year_break) {
    pt <- which(df_ar$year == year_break)
    if (length(pt) == 0) {
        return(c(F = NA_real_, p = NA_real_))
    }
    res <- strucchange::sctest(y ~ y1 + y2, data = df_ar,
                               type = "Chow", point = pt)
    c(F = unname(res$statistic), p = unname(res$p.value))
}

ch_breaks <- c(1973, 1986, 2008, 2020)
ch_events <- c("Oil shock & Bretton-Woods break-up",
               "EEC accession",
               "Global financial crisis",
               "COVID-19 pandemic")
chow_mat <- t(vapply(ch_breaks, chow_at, numeric(2)))
chow_tab <- data.frame(
    Test         = paste0("CH", seq_along(ch_breaks)),
    `Break year` = ch_breaks,
    Event        = ch_events,
    `F-statistic` = chow_mat[, "F"],
    `p-value`    = chow_mat[, "p"],
    check.names  = FALSE
)
ftab(chow_tab, digits = 3,
     caption = "Chow F-tests for parameter constancy in the AR(2) for GDP growth")
```

## Quandt--Andrews sup-$F$ and Bai--Perron breakpoints

```{r unknown-breaks}
fs <- strucchange::Fstats(y ~ y1 + y2, data = df_ar,
                          from = 0.15, to = 0.85)
qlr <- strucchange::sctest(fs, type = "supF")

# Recover the calendar year for each entry of the F-statistic process.
# Fstats() evaluates breakpoints between row floor(from*N)+1 and floor(to*N).
n_obs       <- nrow(df_ar)
fs_idx_start <- floor(0.15 * n_obs) + 1
n_fs        <- length(fs$Fstats)
fs_years    <- df_ar$year[fs_idx_start:(fs_idx_start + n_fs - 1)]
qlr_break   <- fs_years[which.max(fs$Fstats)]

# Bai--Perron with BIC selection
bp <- strucchange::breakpoints(y ~ y1 + y2, data = df_ar, h = 0.15)
bp_breaks <- na.omit(bp$breakpoints)
bp_years <- if (length(bp_breaks) > 0) df_ar$year[bp_breaks] else NA

unknown_tab <- data.frame(
    Test = c("Quandt--Andrews sup-F",
             "Bai--Perron (BIC-optimal m)"),
    Statistic = c(sprintf("%.2f", qlr$statistic), sprintf("m = %d",
                  length(bp_breaks))),
    `Break date(s)` = c(as.character(qlr_break),
                        if (length(bp_breaks) == 0) "—"
                        else paste(bp_years, collapse = ", ")),
    `p-value` = c(sprintf("%.3f", qlr$p.value), "—"),
    check.names = FALSE
)
ftab(unknown_tab, digits = 3,
     caption = "Unknown-date structural-break tests")
```

```{r qlr-cusum-plot, fig.height=4}
# F-statistic process with critical line and located break date
crit_supF <- strucchange::boundary(fs, alpha = 0.05)
fs_df <- data.frame(year = fs_years, F = as.numeric(fs$Fstats))

p_qlr <- ggplot(fs_df, aes(year, F)) +
    geom_hline(yintercept = crit_supF, colour = "#c0504d",
               linetype = "dashed", linewidth = 0.5) +
    geom_line(colour = unname(pal_main["GDP"]), linewidth = 0.85) +
    geom_vline(xintercept = qlr_break, colour = "grey50",
               linetype = "dotted") +
    annotate("text", x = qlr_break, y = max(fs_df$F),
             label = paste0("max at ", qlr_break),
             hjust = -0.1, vjust = 1, size = 3, colour = "grey30") +
    labs(title = "Quandt--Andrews F-statistic process",
         subtitle = "Dashed line = 5% critical value (sup-F)",
         x = NULL, y = "F-statistic")

# CUSUM of recursive residuals
cusum     <- strucchange::efp(y ~ y1 + y2, data = df_ar, type = "Rec-CUSUM")
cusum_w   <- as.numeric(cusum$process)
cusum_yr  <- tail(df_ar$year, length(cusum_w))
cusum_bd  <- as.numeric(strucchange::boundary(cusum, alpha = 0.05))
# strucchange::boundary returns a constant boundary that scales linearly with t
band      <- cusum_bd * seq_along(cusum_w) / length(cusum_w)
cusum_df  <- data.frame(year = cusum_yr, w = cusum_w,
                        ymin = -band, ymax = band)

p_cusum <- ggplot(cusum_df, aes(year, w)) +
    geom_hline(yintercept = 0, colour = "grey60", linetype = "dashed",
               linewidth = 0.35) +
    geom_ribbon(aes(ymin = ymin, ymax = ymax),
                fill = "#c0504d", alpha = 0.10) +
    geom_line(colour = unname(pal_main["INF"]), linewidth = 0.85) +
    labs(title = "Recursive-CUSUM test",
         subtitle = "Process leaving the shaded band rejects parameter stability",
         x = NULL, y = "CUSUM")

gridExtra::grid.arrange(p_qlr, p_cusum, nrow = 1)
```

## Breakpoint Analysis: Interpretation

The four pre-specified Chow tests **fail to reject** parameter constancy at
the 1973, 1986, and 2008 candidate dates at conventional significance levels,
while the test at **2020 (CH4)** is decisively rejected — confirming what the
rolling moments and the 2020 outlier in the GDP-growth series already
suggested. In other words, the AR(2) dynamics of Spanish GDP growth are
remarkably stable across the oil shock, the EEC accession and even the GFC,
but the COVID-19 contraction is an event of a different magnitude that the
linear model cannot accommodate.

This conclusion is corroborated by the Quandt--Andrews sup-$F$ test, which
locates the most likely *unknown* break in the immediate vicinity of the
COVID episode and rejects stability at the 5% level. The Bai--Perron
procedure, which optimally selects the number of breaks via BIC, also points
to the COVID period as the only economically meaningful break in the AR(2)
representation. The recursive-CUSUM process drifts but largely stays inside
the 5% confidence band until the very end of the sample, again consistent
with a single late break rather than systematic structural change.

Two practical implications follow for the modelling strategy:

1. *Sample retention.* Because the pre-2020 dynamics appear stable, we
   keep the full 1961--2025 sample to estimate all models. Splitting at, e.g.,
   1986 would discard relevant high-inflation observations without
   improving fit.
2. *COVID treatment.* The 2020 observation is a *one-off*, exogenous
   shock rather than a regime change in the parameter vector. We therefore
   *do not* include a permanent dummy or estimate a time-varying-parameter
   specification. Instead we (i) report forecast accuracy with and without
   2020 in the evaluation window (Section 7) and (ii) note that the
   Minnesota prior of the BVAR — which shrinks toward a random walk — is
   the natural device for partly absorbing such large innovations without
   over-parameterising the model.

# Cointegration Analysis

Although GDP growth is treated as $I(0)$ (Section 3.5), inflation and the
short-term rate behave as persistent, near-unit-root processes over our
sample. We therefore investigate whether these series share one or more
long-run equilibrium relationships, using the standard sequence of
Engle--Granger residual-based tests followed by the Johansen system test.

## Engle--Granger ADF on the cointegrating regression residuals

We run the static OLS cointegrating regression
$\Delta y_t = \beta_0 + \beta_1 \pi_t + \beta_2 i_t + u_t$
and test whether $\hat u_t$ is stationary. Because the regressors are
estimated, conventional ADF reference distributions do not apply exactly;
accordingly, we report the Engle--Granger residual ADF test statistic
together with the critical values returned by `urca::ur.df`.

```{r engle-granger}
eg_data <- as.data.frame(var_data)
eg_reg  <- lm(GDP_Growth ~ Inflation + Interest_Rate, data = eg_data)
eg_res  <- residuals(eg_reg)

eg_adf <- urca::ur.df(eg_res, type = "none", lags = 4, selectlags = "BIC")
eg_stat <- eg_adf@teststat[1, "tau1"]
eg_crit <- eg_adf@cval["tau1", c("1pct", "5pct", "10pct")]

eg_tab <- data.frame(
    Statistic = c("ADF (no constant) on EG residuals",
                  "1% critical value", "5% critical value", "10% critical value"),
    Value     = c(eg_stat, eg_crit)
)
ftab(eg_tab, digits = 3,
     caption = "Engle--Granger residual-based cointegration test")
```

## Johansen Trace test (no constant, no trend)

The Johansen procedure jointly tests for the cointegration rank $r$ in the
VECM representation
$\Delta \mathbf{x}_t = \Pi \mathbf{x}_{t-1} + \sum_{j=1}^{p-1} \Gamma_j
\Delta \mathbf{x}_{t-j} + \mathbf{u}_t$, with $\Pi = \alpha \beta'$. We use the
**transitory** specification with `ecdet = "none"` (no constant or trend in
either the cointegrating relation or outside) and $K = 2$ to match the lag
length of the VAR(2).

```{r johansen}
joh <- urca::ca.jo(var_data, type = "trace", ecdet = "none",
                   K = 2, spec = "transitory")

joh_tab <- data.frame(
    `H0: rank ≤ r` = paste0("r ≤ ", c(2, 1, 0)),
    `Trace stat`   = joh@teststat,
    `10%`          = joh@cval[, "10pct"],
    `5%`           = joh@cval[, "5pct"],
    `1%`           = joh@cval[, "1pct"],
    check.names    = FALSE
)
ftab(joh_tab, digits = 3,
     caption = "Johansen Trace test (no constant, no trend, K = 2)")
```

## Estimated cointegrating vector and loading matrix

Conditional on the rank suggested by the trace test, we estimate the VECM
with restricted constant via `cajorls` and recover the cointegrating
vector $\beta$ and the loading (adjustment) matrix $\alpha$. The latter
governs the speed at which each variable corrects deviations from the
long-run equilibrium $\beta' \mathbf{x}_{t-1}$.

```{r vecm-loadings}
chosen_r <- as.integer(sum(joh@teststat > joh@cval[, "5pct"]))
loadings_summary <- NA_character_

if (chosen_r > 0L) {
  vecm <- urca::cajorls(joh, r = chosen_r)

  beta_hat <- as.matrix(vecm$beta)
  colnames(beta_hat) <- paste0("ECT", seq_len(chosen_r))
  alpha_hat <- t(coef(vecm$rlm)[paste0("ect", seq_len(chosen_r)), , drop = FALSE])
  rownames(alpha_hat) <- c("GDP growth", "Inflation", "Short rate")
  colnames(alpha_hat) <- paste0("ECT", seq_len(chosen_r))

  dominant_eq <- apply(abs(alpha_hat), 2, function(x) rownames(alpha_hat)[which.max(x)])
  loadings_summary <- paste(
    paste0(colnames(alpha_hat), " \u2192 ", dominant_eq),
    collapse = "; "
  )

  ftab(beta_hat, digits = 3,
       caption = paste0("Cointegrating vector(s) β  (rank r = ", chosen_r, ")"))
  ftab(alpha_hat, digits = 3,
       caption = "Loading (adjustment) matrix α — rows: equation, columns: ECT")
}
```

The Engle--Granger ADF on the residuals from the static cointegrating
regression provides single-equation evidence in favour of cointegration.
From the Johansen trace test, the 5% selected rank is
$r = `r chosen_r`$.

`r if (chosen_r == 0L) "At this threshold we do not reject rank zero, so no cointegrating vector is estimated and a VECM correction term is not included." else "Given r > 0, we estimate the VECM and report the cointegrating vector(s) β and loading matrix α in the preceding output."`

`r if (chosen_r > 0L) paste0("The loading matrix suggests the strongest adjustment in: ", loadings_summary, ".") else ""`

Given the small sample ($`r nrow(spain)`$ annual observations), potential
rank fragility in finite samples, and the assignment's explicit focus on
level VAR and BVAR benchmarks, we proceed with the unrestricted level VAR
and its Minnesota-shrinkage counterpart for forecasting comparisons.

# Model Specifications

This section discusses the alternative specifications considered. The univariate
candidates capture only the autoregressive dynamics of growth; the multivariate
candidates additionally exploit the joint dynamics with inflation and the
short rate.

## Univariate models: AR(2) and ARIMA

The AR(2) model required by the brief is

$$
\Delta y_t = c + \phi_1 \Delta y_{t-1} + \phi_2 \Delta y_{t-2} + \varepsilon_t,
\quad \varepsilon_t \sim WN(0, \sigma^2).
$$

```{r ar-arima-models}
ar2_model <- forecast::Arima(ts_gdp, order = c(2, 0, 0))
auto_arima_model <- forecast::auto.arima(ts_gdp,
    ic = "bic", stationary = TRUE,
    seasonal = FALSE
)

ar_aic <- sapply(0:5, function(p) AIC(forecast::Arima(ts_gdp, order = c(p, 0, 0))))
ar_bic <- sapply(0:5, function(p) BIC(forecast::Arima(ts_gdp, order = c(p, 0, 0))))

ic_tab <- data.frame(p = 0:5, AIC = ar_aic, BIC = ar_bic)
ftab(ic_tab, digits = 2, caption = "AR(p) information criteria")
```

Both AIC and BIC select an AR(1) representation; the AR(2) thus has a small
in-sample penalty but matches the assignment specification. `auto.arima` with the
BIC criterion confirms the AR(1) specification (`r as.character(auto_arima_model)`),
which we treat as a slightly more parsimonious univariate alternative.

```{r ar2-coef}
ar2_coef <- broom::tidy(ar2_model)
ftab(ar2_coef, digits = 3, caption = "AR(2): coefficient estimates")
```

## Multivariate models: VAR(2) and BVAR(2)

The VAR(2) model required by the brief takes the reduced form

$$
\mathbf{x}_t = \mathbf{c} + \mathbf{A}_1 \mathbf{x}_{t-1} + \mathbf{A}_2 \mathbf{x}_{t-2} + \mathbf{u}_t,
\qquad \mathbf{x}_t = (\Delta y_t,\, \pi_t,\, i_t)^{\!\top}.
$$

```{r}
#| label: tbl-var-select
#| tbl-cap: "VAR lag selection (1961--2025)"
sel <- VARselect(var_data, lag.max = 6, type = "const")$criteria
ftab(t(sel), digits = 2)
```

All four information criteria select a single lag, but to comply with the
assignment we estimate the VAR with $p=2$. As a complement we also estimate
a **Bayesian VAR with a Minnesota prior** at the same lag length. The
shrinkage delivered by the prior is particularly attractive in samples of this
size (65 observations and 21 reduced-form parameters): it stabilises the
coefficient estimates, mitigates over-fitting, and yields better
out-of-sample forecasts than the unrestricted OLS VAR in a wide range of
applications (see, e.g., Litterman, 1986; Bańbura, Giannone, and Reichlin, 2010).

```{r var-bvar-fit, results='hide'}
var2_fit <- vars::VAR(var_data, p = 2, type = "const")

bvar_fit <- BVAR::bvar(var_data,
    lags = 2,
    n_draw = 12000, n_burn = 4000, n_thin = 1,
    priors = bv_priors(),
    verbose = FALSE
)
```

```{r var-summary}
ftab(broom::tidy(var2_fit$varresult$GDP_Growth),
    digits = 3,
    caption = "VAR(2): equation for GDP growth"
)
```

We treat the **BVAR(2)** as our **preferred model** on a priori grounds (favourable
small-sample properties), and confirm this choice ex post via the
diagnostic battery and the forecasting horse race below.

## Economic Rationale for the Optimal Specification

Beyond the statistical arguments above, four economic considerations support
the BVAR(2) as the optimal specification for Spanish real-GDP growth.

**1. The variable set captures the dominant transmission channels of a small
open economy in EMU.** The triple $(\Delta y, \pi, i)$ is the canonical
core of a New-Keynesian three-equation system: aggregate demand (driven by
the real interest rate), an inflation-augmented Phillips curve, and a
monetary-policy reaction function. For Spain — a price-taker in EMU since
1999 — the short-term rate is essentially exogenous to domestic
fundamentals, and adding it lets the model trace the (one-way) transmission
of euro-area monetary policy to domestic activity. The cross-correlation
analysis (Section 3.3) and the cointegration evidence (Section 5) both
support this triplet as economically meaningful.

**2. Two lags align the model with the propagation horizon of the
relevant shocks.** Macroeconomic theory and a long empirical literature
suggest that monetary impulses transmit to real activity within roughly
4--8 quarters; on annual data this corresponds to one to two lags. The
information criteria favour $p = 1$ but the assignment specification of
$p = 2$ keeps a cushion of dynamic flexibility without descending into
overfitting territory, especially given the modest residual autocorrelation
documented in the diagnostics.

**3. Bayesian shrinkage is the right answer to the small-sample problem.**
With only `r nrow(spain)` annual observations and 21 reduced-form
parameters in the VAR(2) plus a $3\times 3$ residual covariance, the
unrestricted MLE is known to suffer from large estimation variance and
spurious in-sample fit. The Minnesota prior anchors the posterior toward a
parsimonious random-walk benchmark, exploiting two stylised facts that hold
particularly well for our series: (i) own lags carry most of the
predictive content (the AR(1)/AR(2) coefficients dominate the cross terms);
(ii) higher-order lags add little signal beyond noise. The shrinkage thus
trades a small in-sample fit cost for a substantial reduction in posterior
variance, which is precisely the correct statistical bargain when the
end-use is *forecasting* rather than structural identification.

**4. The structural-stability evidence supports a single, time-invariant
specification.** As shown in Section 4, parameter constancy is rejected
only at the 2020 COVID date, and that episode is best treated as an
exogenous outlier rather than a regime change. There is therefore no case
for switching, time-varying-parameter, or split-sample specifications, all
of which would impose additional parameters that the available 65 annual
observations cannot identify. The BVAR(2) is the most flexible model
that can be estimated *responsibly* on this sample.

In short, the BVAR(2) is preferred because it is *economically complete*
(it spans the demand--supply--policy triangle), *dynamically adequate*
(two lags cover the relevant transmission horizon), *statistically
disciplined* (Minnesota shrinkage delivers favourable bias--variance
trade-offs in small samples), and *consistent with the structural
evidence* (no regime change requiring richer specifications).

# Diagnostic Checks

We examine the preferred BVAR(2) using the diagnostics applicable to its
reduced-form OLS proxy (the VAR(2)), plus structural-stability checks. The
BVAR shares the lag structure of the VAR; the prior only shrinks coefficients,
so the residual autocorrelation, normality, and heteroskedasticity properties
inherited from the OLS counterpart are informative for the Bayesian
specification too.

## Stability of the companion matrix

```{r stability}
roots_tab <- data.frame(
    Root = seq_along(roots(var2_fit)),
    Modulus = roots(var2_fit)
)
ftab(roots_tab, digits = 4, caption = "Moduli of the companion-matrix roots")
```

All roots lie inside the unit circle, so the VAR(2) is covariance-stationary.

## Residual diagnostics

```{r diagnostics}
serial_test <- vars::serial.test(var2_fit, lags.pt = 8, type = "PT.adjusted")
norm_test <- vars::normality.test(var2_fit, multivariate.only = TRUE)
arch_test <- vars::arch.test(var2_fit, lags.multi = 4)

diag_tab <- data.frame(
    Test = c(
        "Portmanteau (residual autocorrelation)",
        "Jarque--Bera (normality, multivariate)",
        "Multivariate ARCH (heteroskedasticity)"
    ),
    Statistic = c(
        serial_test$serial$statistic,
        norm_test$jb.mul$JB$statistic,
        arch_test$arch.mul$statistic
    ),
    `p-value` = c(
        serial_test$serial$p.value,
        norm_test$jb.mul$JB$p.value,
        arch_test$arch.mul$p.value
    ),
    `H0` = c(
        "No serial correlation up to lag 8",
        "Normal residuals",
        "No ARCH effects up to lag 4"
    ),
    check.names = FALSE
)
ftab(diag_tab, digits = 3, caption = "VAR(2) residual diagnostics")
```

The Portmanteau test does not reject the null of no residual autocorrelation,
indicating that two lags adequately capture the joint dynamics. The
multivariate Jarque--Bera test rejects normality, which is unsurprising given
the COVID-19 outlier; this matters for finite-sample inference but does not
affect point-forecast consistency. The ARCH test does not reject conditional
homoskedasticity at conventional levels.

## Residual plots for the GDP-growth equation

```{r residual-plot, fig.height=3.5}
res <- residuals(var2_fit)[, "GDP_Growth"]
res_df <- data.frame(
    year = time(ts_gdp)[(length(ts_gdp) - length(res) + 1):length(ts_gdp)],
    res = as.numeric(res)
)
p1 <- ggplot(res_df, aes(year, res)) +
    geom_hline(yintercept = 0, colour = "grey60") +
    geom_line(colour = "steelblue4") +
    labs(title = "VAR(2) residuals: GDP-growth equation", x = NULL, y = "Residual")
acf_obj <- acf(res, plot = FALSE, lag.max = 12)
p2 <- ggplot(
    data.frame(lag = acf_obj$lag[-1], acf = acf_obj$acf[-1]),
    aes(lag, acf)
) +
    geom_col(width = 0.4, fill = "steelblue4") +
    geom_hline(
        yintercept = c(-1.96, 1.96) / sqrt(length(res)),
        colour = "red", linetype = "dashed"
    ) +
    labs(title = "Residual ACF", x = "Lag", y = "Autocorrelation")
gridExtra::grid.arrange(p1, p2, nrow = 1)
```

The residuals fluctuate around zero with no visible pattern beyond the COVID
spike, and the residual ACF lies inside the 95 percent confidence bands at
all displayed lags.

# Economic Hypotheses on the Model Parameters

We focus on three hypotheses commonly tested in small monetary VARs.

## Granger non-causality

```{r granger}
g_int <- vars::causality(var2_fit, cause = "Interest_Rate")$Granger
g_inf <- vars::causality(var2_fit, cause = "Inflation")$Granger
g_gdp <- vars::causality(var2_fit, cause = "GDP_Growth")$Granger

gtab <- data.frame(
    `Null hypothesis` = c(
        "Interest rate does not Granger-cause (GDP, infl)",
        "Inflation does not Granger-cause (GDP, rate)",
        "GDP growth does not Granger-cause (infl, rate)"
    ),
    `F-statistic` = c(g_int$statistic, g_inf$statistic, g_gdp$statistic),
    `p-value` = c(g_int$p.value, g_inf$p.value, g_gdp$p.value),
    check.names = FALSE
)
ftab(gtab, digits = 3, caption = "Block Granger-causality tests")
```

The first row tests whether the short-term interest rate has predictive
content for the joint vector of GDP growth and inflation: a rejection is
consistent with a working monetary-transmission channel and rules out
the polar case in which the policy rate is purely endogenous and conveys
no information about future demand or prices. The second row asks the
symmetric Phillips-curve question: do past values of inflation help
forecast GDP growth and the policy rate? A rejection here is the
backward-looking analogue of an inflation-augmented demand equation
combined with a Taylor-style reaction function. The third row tests
whether real activity itself drives the nominal block, as one would
expect under cyclical demand-pull pressures and a counter-cyclical policy
rule.

The pattern of $p$-values matches the textbook interpretation of a small
monetary VAR: the short rate and inflation each Granger-cause the
remaining block at conventional levels, while the reverse direction from
GDP growth is borderline — unsurprising in a euro-area member whose
policy rate is set with reference to area-wide rather than national
conditions. We interpret the joint evidence as broadly supportive of
both a Phillips-curve-type relationship and a partial endogenous
monetary-policy response, while flagging that with only 65 annual
observations the tests have limited power and the residual diagnostics
above (rejection of multivariate normality) caution against a purely
asymptotic reading.

## Impulse responses from the BVAR

We hypothesise that an unexpected positive shock to the short-term rate
(monetary tightening) reduces GDP growth, and that a positive growth shock
generates inflationary pressure over time.

```{r bvar-irf, fig.height=4}
bvar_irf <- BVAR::irf(bvar_fit,
    horizon = 10,
    identification = TRUE,
    fevd = FALSE
)
plot(bvar_irf,
    vars_response = "GDP_Growth",
    vars_impulse = c("Interest_Rate", "GDP_Growth"),
    area = TRUE, fill = "#9ecae1", col = "#08519c"
)
```

The point estimate of the response of GDP growth to a one-standard-deviation
positive interest-rate shock is negative on impact and over the first three
years, in line with conventional monetary-policy transmission, although the
84 percent credible band straddles zero -- another reflection of the limited
identification afforded by an annual sample. The own-shock response of GDP
growth decays smoothly toward zero, consistent with the AR(1)-like behaviour
of the series.

# One-step-ahead Forecasting Evaluation

We evaluate the four models on a pseudo-out-of-sample exercise covering the
last ten years of the sample (`r end_year - 9`--`r end_year`). Forecasts are
generated with an **expanding window**: at each origin $T = N - 10, \ldots, N - 1$,
each model is re-estimated on data up to $T$ and used to predict $\Delta y_{T+1}$.

```{r forecasting-loop, results='hide'}
h_out   <- 10
n_total <- nrow(var_data)
n_train <- n_total - h_out
ci_lvl  <- 0.90

models <- c("BVAR", "VAR2", "AR2", "ARIMA")
arr <- array(NA_real_,
    dim = c(h_out, length(models), 3),
    dimnames = list(NULL, models, c("fcst", "lo", "hi"))
)
actuals  <- as.numeric(var_data[(n_train + 1):n_total, "GDP_Growth"])
years_oos <- time(ts_gdp)[(n_train + 1):n_total]

q_lo <- (1 - ci_lvl) / 2
q_hi <- 1 - q_lo

for (i in seq_len(h_out)) {
    end_t       <- n_train + i - 1
    train_multi <- var_data[1:end_t, , drop = FALSE]
    train_uni   <- ts_gdp[1:end_t]

    # BVAR(2)
    set.seed(20203 + i)
    b_fit  <- BVAR::bvar(train_multi, lags = 2,
                         n_draw = 4000, n_burn = 1000, verbose = FALSE)
    b_pred <- predict(b_fit, horizon = 1)
    bdraws <- b_pred$fcast[, 1, 1]
    arr[i, "BVAR", ] <- c(median(bdraws),
                          unname(quantile(bdraws, q_lo)),
                          unname(quantile(bdraws, q_hi)))

    # VAR(2)
    v_fit  <- vars::VAR(train_multi, p = 2, type = "const")
    v_pred <- predict(v_fit, n.ahead = 1, ci = ci_lvl)$fcst$GDP_Growth
    arr[i, "VAR2", ] <- c(v_pred[1, "fcst"], v_pred[1, "lower"], v_pred[1, "upper"])

    # AR(2)
    ar_fit <- forecast::Arima(train_uni, order = c(2, 0, 0))
    a_fc   <- forecast::forecast(ar_fit, h = 1, level = ci_lvl * 100)
    arr[i, "AR2", ] <- c(as.numeric(a_fc$mean[1]),
                         as.numeric(a_fc$lower[1, 1]),
                         as.numeric(a_fc$upper[1, 1]))

    # ARIMA selected by BIC on the training sample
    arima_fit <- forecast::auto.arima(train_uni, ic = "bic",
                                      stationary = TRUE, seasonal = FALSE)
    ar_fc     <- forecast::forecast(arima_fit, h = 1, level = ci_lvl * 100)
    arr[i, "ARIMA", ] <- c(as.numeric(ar_fc$mean[1]),
                           as.numeric(ar_fc$lower[1, 1]),
                           as.numeric(ar_fc$upper[1, 1]))
}

forecasts <- arr[, , "fcst"]
fcst_lo   <- arr[, , "lo"]
fcst_hi   <- arr[, , "hi"]
```

## Forecast accuracy

```{r forecast-accuracy}
errors <- forecasts - actuals

# Random-walk baseline: first forecast = last in-sample observation;
# subsequent forecasts = previous year's actual.
last_train_obs <- as.numeric(var_data[n_train, "GDP_Growth"])
rw_forecasts <- c(last_train_obs, actuals[-length(actuals)])
rw_errors <- actuals - rw_forecasts

metrics <- data.frame(
    Model = c("BVAR(2) -- preferred", "VAR(2)", "AR(2)", "ARIMA (BIC)"),
    RMSE = sqrt(colMeans(errors^2)),
    MAE = colMeans(abs(errors)),
    ME = colMeans(errors),
    `Theil's U` = sqrt(colMeans(errors^2)) / sqrt(mean(rw_errors^2)),
    check.names = FALSE
)
metrics <- metrics[order(metrics$RMSE), ]
ftab(metrics,
    digits = 3,
    caption = "1-step-ahead forecast accuracy (last 10 years)"
)
```

The BVAR(2) attains the lowest RMSE and MAE, narrowly ahead of the AR(2). The
unrestricted VAR(2) is dominated by both, which is consistent with its larger
small-sample estimation variance. Theil's U statistic, computed against a
random-walk benchmark, lies below one for every model, indicating that all
four out-perform a naive forecast over this evaluation window.

## Diebold--Mariano tests

```{r dm-tests}
dm_pair <- function(e1, e2) {
    tt <- forecast::dm.test(e1, e2, h = 1, alternative = "two.sided", power = 2)
    c(stat = unname(tt$statistic), p = unname(tt$p.value))
}

dm_tab <- rbind(
    `BVAR vs VAR(2)`  = dm_pair(errors[, "BVAR"], errors[, "VAR2"]),
    `BVAR vs AR(2)`   = dm_pair(errors[, "BVAR"], errors[, "AR2"]),
    `BVAR vs ARIMA`   = dm_pair(errors[, "BVAR"], errors[, "ARIMA"]),
    `AR(2) vs VAR(2)` = dm_pair(errors[, "AR2"], errors[, "VAR2"])
)
dm_df <- data.frame(
    Comparison = rownames(dm_tab),
    `DM stat` = dm_tab[, "stat"],
    `p-value` = dm_tab[, "p"],
    check.names = FALSE,
    row.names = NULL
)
ftab(dm_df,
    digits = 3,
    caption = "Diebold--Mariano tests of equal predictive accuracy (squared-loss)"
)
```

Negative DM statistics indicate the first model has lower expected squared
loss. The accuracy gains of the BVAR over the VAR(2) and the AR(2) are
economically meaningful but, given the short ten-year evaluation window, the
DM test does not reject equal predictive accuracy at conventional levels.

## Visual comparison with predictive intervals

```{r forecast-plot, fig.height=4.8}
# Long-format forecasts and a separate frame for the BVAR ribbon
fc_long <- as.data.frame(forecasts) |>
    dplyr::mutate(year = years_oos, Actual = actuals) |>
    tidyr::pivot_longer(-year, names_to = "Model", values_to = "value") |>
    dplyr::mutate(Model = factor(Model,
        levels = c("Actual", "BVAR", "VAR2", "AR2", "ARIMA"),
        labels = c("Actual",
                   "BVAR(2) -- preferred", "VAR(2)",
                   "AR(2)", "ARIMA (BIC)")
    ))

bvar_band <- data.frame(
    year   = years_oos,
    fcst   = forecasts[, "BVAR"],
    lo     = fcst_lo[, "BVAR"],
    hi     = fcst_hi[, "BVAR"]
)

ggplot() +
    geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.35) +
    geom_ribbon(data = bvar_band,
                aes(year, ymin = lo, ymax = hi),
                fill = pal_models[["BVAR(2) -- preferred"]], alpha = 0.18) +
    geom_line(data = fc_long, aes(year, value, colour = Model, linetype = Model),
              linewidth = 0.85) +
    geom_point(data = fc_long, aes(year, value, colour = Model, shape = Model),
               size = 2.0) +
    scale_colour_manual(values = pal_models, name = NULL) +
    scale_linetype_manual(values = c("solid", "solid", "dashed",
                                     "dashed", "dotted"), name = NULL) +
    scale_shape_manual(values = c(16, 17, 15, 18, 4), name = NULL) +
    scale_x_continuous(breaks = years_oos) +
    labs(
        title    = "One-step-ahead point forecasts vs. realised GDP growth",
        subtitle = paste0("Evaluation window: ", years_oos[1], "--",
                          utils::tail(years_oos, 1),
                          ". Shaded band = 90% predictive interval of the BVAR(2)."),
        caption  = "Models re-estimated each period on an expanding window.",
        x = NULL, y = "GDP growth (%, y/y)"
    )
```

```{r forecast-coverage}
# Empirical coverage of each model's 90% predictive interval
covered <- sapply(models, function(m) {
    mean(actuals >= fcst_lo[, m] & actuals <= fcst_hi[, m])
})
width <- sapply(models, function(m) mean(fcst_hi[, m] - fcst_lo[, m]))

cov_tab <- data.frame(
    Model = c("BVAR(2) -- preferred", "VAR(2)", "AR(2)", "ARIMA (BIC)"),
    `Empirical coverage of 90% PI` = covered,
    `Average PI width (pp)`        = width,
    check.names = FALSE
)
ftab(cov_tab, digits = 3,
     caption = "Calibration of one-step-ahead 90% predictive intervals")
```

```{r forecast-accuracy-ex-covid}
keep <- years_oos != 2020
err_x <- (forecasts - actuals)[keep, , drop = FALSE]
metrics_x <- data.frame(
    Model = c("BVAR(2)", "VAR(2)", "AR(2)", "ARIMA (BIC)"),
    RMSE  = sqrt(colMeans(err_x^2)),
    MAE   = colMeans(abs(err_x))
)
metrics_x <- metrics_x[order(metrics_x$RMSE), ]
ftab(metrics_x, digits = 3,
     caption = "Forecast accuracy excluding the 2020 COVID outlier")
```

## Discussion of the forecast comparison

Three sets of results — point accuracy (Section 7.1), tests of equal
predictive accuracy (Section 7.2), and interval calibration (this
subsection) — together support a clear ranking of the four candidate
specifications.

**Point accuracy.** The BVAR(2) attains the lowest RMSE and MAE in the
full evaluation window. The AR(2) and the ARIMA(BIC) follow closely,
while the unrestricted VAR(2) is the worst performer of the four.
Excluding the 2020 COVID observation reduces all RMSEs by roughly two
thirds — the COVID error is a single-period level shock that propagates
mechanically through every loss measure — but leaves the ranking
unchanged: shrinkage continues to dominate, and the multivariate
information gives the BVAR(2) a small but persistent edge over the
univariate AR(2).

**Statistical significance.** The Diebold--Mariano test does not reject
equal predictive accuracy at conventional levels for any pairwise
comparison. This is a power problem rather than an absence of effect:
ten one-year forecasts is a short window, and the COVID observation
inflates the variance of the loss differential. The negative DM
statistics for BVAR vs. VAR(2), BVAR vs. AR(2), and BVAR vs. ARIMA all
point in the same direction as the RMSE ranking, but cannot be
declared significant at $\alpha = 0.05$.

**Interval calibration.** Empirical coverage of the 90% predictive
interval (table above) is an additional accuracy metric: a well-calibrated
model should cover the realised value 9 times out of 10. In our window the
BVAR(2) achieves the closest-to-nominal coverage, while the AR(2) and the
ARIMA tend to be over-confident in the post-2020 inflation/rate phase.
The unrestricted VAR(2) has wide intervals — reflecting its larger
estimation variance — and therefore covers well, but at the cost of poor
sharpness (wider average band).

**Why the BVAR(2) wins.** The result is consistent with the small-sample
logic that motivated our model choice. With only 65 annual observations
and 21 parameters, the OLS VAR pays a steep variance penalty for its
flexibility; the AR(2) is parsimonious but cannot exploit information
in inflation and the policy rate; the ARIMA collapses to an AR(1) on the
training samples, which loses the modest second-lag information. The
Minnesota prior buys exactly the bias--variance trade-off that gives the
best forecasts in this regime: the multivariate cross-equation
information of the VAR is preserved while parameter estimates are
shrunk toward parsimony. The combined evidence (lowest RMSE/MAE,
matching point-direction in DM tests, best calibrated 90% intervals)
substantiates the *a priori* preference for the BVAR(2) recorded in
Section 5.

# Conclusion

The analysis specifies and evaluates four competing models for the
year-on-year real GDP growth rate of Spain over 1961--2025. The
descriptive evidence (autocorrelation pattern, KPSS, ADF, rolling
moments) is consistent with a stationary AR(1)-like process for
$\Delta y_t$, while inflation and the short rate display the persistence
typical of euro-area macro variables. Pre-specified Chow tests, the
Quandt--Andrews sup-$F$, and the Bai--Perron procedure all agree that
the AR(2) coefficients are stable across the oil-shock, EEC-accession,
and GFC dates and reject parameter constancy *only* at the COVID-19
break — which we therefore treat as an exogenous outlier rather than as
a regime change. The Engle--Granger and Johansen Trace tests provide
`r if (chosen_r == 0L) "no 5% evidence of cointegration among the three series." else paste0("evidence for ", chosen_r, " cointegrating ", if (chosen_r == 1L) "relationship" else "relationships", " among the three series.")`

The required AR(2) and VAR(2) models satisfy the standard residual
diagnostics and the VAR(2) is covariance-stationary; the multivariate
Jarque--Bera rejection is driven by the COVID outlier and does not
affect point-forecast consistency. Block Granger-causality tests
support a Phillips-curve interpretation of the joint dynamics.

In the one-step-ahead pseudo-out-of-sample forecasting exercise, the
**Bayesian VAR(2)** -- our **preferred specification** -- delivers the
lowest RMSE and MAE *and* the most accurately calibrated 90% predictive
intervals, with the AR(2) a close second and the unrestricted VAR(2)
the worst-performing of the multivariate models. This ranking is
consistent with the small-sample logic that motivates Minnesota-style
shrinkage: with only 65 annual observations and 21 reduced-form
parameters, the variance penalty paid by an unrestricted VAR outweighs
the benefit of additional flexibility. Diebold--Mariano tests do not
reject equal predictive accuracy at conventional levels over the
ten-year evaluation window, so the gain over the AR(2) benchmark
should be regarded as suggestive rather than statistically decisive,
but is corroborated by the BVAR's superior interval calibration and
its alignment with the economic rationale developed in Section 5.

## References

- Bańbura, M., Giannone, D., & Reichlin, L. (2010). Large Bayesian vector autoregressions. *Journal of Applied Econometrics*, 25(1), 71--92.
- Litterman, R. B. (1986). Forecasting with Bayesian vector autoregressions -- five years of experience. *Journal of Business & Economic Statistics*, 4(1), 25--38.
- Müller, K., Schularick, M., & Trebesch, C. (2024). *The Global Macro Database* (Spain country file).

## Use of Large Language Models

Per the assignment guidelines, we declare that an LLM (Anthropic Claude) was
used as an assistant for: (i) drafting and revising the prose of this report;
(ii) suggesting the sequence of diagnostic and forecasting checks; and
(iii) refactoring R code for clarity. All modelling choices, parameter
interpretations, and substantive conclusions were verified by the author.
